The population of long-period exoplanets --- \\ Fully-automated discovery \& characterization of long-period transiting exoplanet candidates in the \kepler\ archive

### 2016-04-29 document/ms.tex
moving GP to appendix

% no more bad lines!
\sloppy\sloppypar

@@

The transit method of exoplanet detection and characterization has been
demonstrated as a powerful method for building systematic catalogs of
exoplanets.
Despite the great success of the \kepler\ Mission with thousands of planet
discoveries \citep{Burke:2014, Rowe:2015}, current methods for exoplanet
discoveries are currently limited in the range of orbital periods that can be
studied.
Specifically, the standard transit search procedures only discover signals
with at least three observed transits \citep[for example][]{Petigura:2013,
Burke:2014, Rowe:2015}.
For \kepler, with a baseline of about four years, this sets an absolute upper
limit on the detectable periods of about two years.
In the Solar System, Jupiter---with a period of 12 years---dominates the
planetary dynamics and, since it would only exhibit at most one transit in the
\kepler\ data, it would be missed by most existing transit search procedures.
It is possible to discover long-period planets like this using targeted radial
velocity (RV) surveys \citep[for example][]{Butler:2006, Knutson:2014}
\dfmtodo{CITE Bryan \etal} but the cost of implementing a systematic RV search
is substantially higher than searching the existing and forthcoming
photometric data for single transits.

There are two main technical barriers to a search for single transit events.
The first is that the transit probability for long-period planets is very low;
scaling as $\propto P^{-5/3}$ for orbital periods longer than the
baseline of contiguous observations.
Therefore, even if long-period planets are intrinsically common, they will
still be underrepresented in a transiting sample.
The second challenge is that there are many signals in the observed
light curves caused by stochastic processes---both instrumental
(pointing jitter, temperature variations, \etc) and astrophysical (stellar
variability, \etc)---that can masquerade as transits.
Even using the most sophisticated methods for removing this variability, false
signal far outnumber the true single transits.

@@

To study the occurrence rate of long-period exoplanets, we have developed a
fully-automated method to search the archival \kepler\ light curves and
reliably discover long-period planets with only one or two transits.
Using this method, we search the 40,000 brightest Sun-like stars and detect
XXX long-period astrophysical transit candidates.
Of these, YYY are likely to be exoplanets with radii equal to or smaller than
Jupiter's radius.
Using transit injection and recovery, we measure the completeness of our
search procedure.
We estimate that the population level astrophysical false positive rate in
this candidate sample is about ZZZ\%.
By injecting synthetic transit signals into the \kepler\ light curves, we
measure the completeness of our search method.
Based on this catalog of exoplanet candidates and the search completeness, we
constrain the occurrence rate of long-period exoplanets.
For large planets, these results are consistent with previous measurements of
long-period, giant planet occurrence rates based on radial velocity surveys
and we demonstrate that small---and possibly rocky---planets are more common
than their larger counterparts even at long orbital periods.

@@

The model comparison step (step 2) is the key component of our method that
enables robust automation but it is also computationally expensive because we
must estimate the marginalized likelihoods for the light curve under several
different models describing a transit and many other processes that ``look''
like transits but are actually caused by noise.
This step is also conservative; unless a signal is a very convincing transit,
it won't pass the test.
In practice, this means that all but the highest signal-to-noise events will
be rejected at this step.
Therefore, in the inexpensive first step---the initial candidate search---we
can restrict the candidate list to high signal-to-noise events without a
substantial loss in detection efficiency.

\subsection{Finding box-shaped candidate events}

It is not computationally feasible to run a full model comparison at every
time in the light curve so we must first find potentially interesting events.
For our purposes, ``interesting'' means high signal-to-noise and previously
unknown.
We restrict our search to high signal-to-noise transits since the

@@

As previously mentioned, the following steps will only detect high
signal-to-noise transits.
Therefore, we use this
To avoid edge effects, we apodize this detection scalar near any large gaps in
the time series using a logistic function with width equal to one transit
duration.
Finally, we estimate the noise in the time series using a robust running
windowed variance estimate of the detection scalar and accepting peaks more
than 25-times this noise as candidates.

@@

In this step of the method, the goal is to discard any signals that are not
sufficiently ``transit-like'' in shape.
To quantify this decision, we perform a model comparison between a physical
transit model and a set of other parameterized models for systematics.
In order for a candidate to pass this vetting step, the transit model must be
``preferred'' to any other model as measured using the Bayesian Information
Criterion (BIC).
The BIC is not the optimal choice for this model comparison but it is
computationally intractable to compute thousands of precise marginalized
likelihoods for each model.
The BIC can be efficiently computed and it exhibits the desired
behavior---increasing with the likelihood but flexible models are
penalized---and we find that it performs sufficiently well in practice.

@@

For each model, we describe the data using a Gaussian Process (GP;
\dfmtodo{cite}) with a Mat\'ern-3/2 covariance and mean given by the chosen
model $m_k(t;\,\meanpars)$ parameterized by the parameter vector \meanpars.
This generative model is specified by the following likelihood function:
\begin{eqnarray}
\mathcal{L} = \ln p_k(\bvec{y}\,|\,\meanpars,\,\kernpars) &=&
- \frac{1}{2}\,\bvec{r}_k(\meanpars)^\T\,K(\kernpars)^{-1}\,
    \bvec{r}_k(\meanpars)
- \frac{1}{2}\log\det K(\kernpars) - \frac{N}{2} \log{2\,\pi}
\end{eqnarray}
where \bvec{y} is the vector of flux measurements, \bvec{t} are the
corresponding times, and
\begin{eqnarray}
\bvec{r}_k(\meanpars) &=& \bvec{y} - m_k(\bvec{t};\,\meanpars) \quad.
\end{eqnarray}
The covariance matrix $K(\kernpars)$ is parameterized by the parameters
$\kernpars = (\alpha,\,\tau)$ and its elements are given by
\begin{eqnarray}
\left[ K(\kernpars) \right]_{ij} &=& \sigma_i^2\,\delta_{ij}
    + \alpha^2 \left[ 1+\frac{|t_i - t_j|}{\sqrt{3}\,\tau} \right]
      \exp \left(-\frac{|t_i - t_j|}{\sqrt{3}\,\tau}\right)
\end{eqnarray}
where $\sigma_i$ is the reported uncertainty on the $i$-th flux measurement.

The gradients of the likelihood with respect to the parameters \meanpars\
and \kernpars\ are given by
\begin{eqnarray}
\frac{\dd\ln p_k(\bvec{y}\,|\,\meanpars,\,\kernpars)}{\dd \meanpars} &=&
\frac{\dd m_k(\bvec{t};\,\meanpars)}{\dd\meanpars}^\T \, K(\kernpars)^{-1} \,
    m_k(\bvec{t};\,\meanpars)
\end{eqnarray}
and
\begin{eqnarray}
\frac{\dd\ln p_k(\bvec{y}\,|\,\meanpars,\,\kernpars)}{\dd \kernpars} &=&
\frac{1}{2}\,\mathrm{Tr}\left(
    \left[ \bvec{\phi}\,\bvec{\phi}^\T - K(\kernpars)^{-1} \right]
    \,\frac{\dd K(\kernpars)}{\dd\kernpars}
\right)
\end{eqnarray}
where
\begin{eqnarray}
\bvec{\phi} &=& K(\kernpars)^{-1}\,\bvec{r}_k(\meanpars) \quad.
\end{eqnarray}
Therefore, if we can compute the gradient of the model with respect to its
parameters, we can efficiently optimize this function to find the maximum
likelihood parameters using a gradient-based non-linear optimization method.
To compute the gradients of the transit model, we use a compile-time automatic
differentiation library \dfmtodo{cite Ceres} and for the other models, we
choose models that can be easily differentiated analytically.

@@

The goal of this project is to discover the transits of long-period planets
that have not yet been discovered.
Therefore, when studying the light curve of an eclipsing binary star or a star
with known transiting planet candidates---on shorter periods---we also remove
all the in-transit data for the candidate using the parameters provided by the
\project{NASA Exoplanet
Archive}\footnote{\url{http://exoplanetarchive.ipac.caltech.edu/}; We
downloaded the \texttt{cumulative} table of \kepler\ Objects of Interest on
2015-03-25.}.

@@

To measure the completeness of our long-period transit search method, we
exploit the fact that transit signals are sparse and rare.
Therefore, most light curves contain no transits and we can reliably measure
the recovery rate of our method on synthetic transit signals---with known
properties---injected into real light curves.
This procedure is standard practice in the transit literature and it has been
used to determine the completeness of the KOI catalog (\dfmtodo{cite}) and
other independent transit searches (\dfmtodo{cite}).

@@

The discovery and characterization of transiting planets based on a single
transit event is crucial for the future of transiting exoplanet surveys.
Many of the most dynamically influential planets---like Jupiter in our Solar
system---exhibit only a single transit in the full observational baseline of
the \kepler.
This will become even more of a problem as upcoming surveys move to shorter
contiguous observations.
For example, the \tess\ Mission is planned to get full-sky coverage at
half-hour cadence but most of the sky will only be targeted for a month.
This means that even habitable zone planets orbiting cool stars will transit
their host \emph{at most once in the entire lifetime of the Mission}!

@@

\section{Supervised classification}

>>>>

% no more bad lines!
\sloppy\sloppypar
\widowpenalty10000
\clubpenalty500

@@

The transit method of exoplanet detection and characterization has been
demonstrated as a powerful method for building systematic catalogs of
exoplanets.
Despite the great success of the \kepler\ Mission with thousands of planet
discoveries \citep{Burke:2014, Rowe:2015}, current methods for exoplanet
discoveries are currently limited in the range of orbital periods that can be
studied.
Specifically, the standard transit search procedures only discover signals
with at least three observed transits \citep[for example][]{Petigura:2013,
Burke:2014, Rowe:2015}.
For \kepler, with a baseline of about four years, this sets an absolute upper
limit on the detectable periods of about two years.
In the Solar System, Jupiter~--~with a period of 12 years~--~dominates the
planetary dynamics and, since it would only exhibit at most one transit in the
\kepler\ data, it would be missed by most existing transit search procedures.
It is possible to discover long-period planets like this using targeted radial
velocity (RV) surveys \citep[for example][]{Butler:2006, Knutson:2014}
\dfmtodo{CITE Bryan \etal} but the cost of implementing a systematic RV search
is substantially higher than searching the existing and forthcoming
photometric data for single transits.

There are two main technical barriers to a search for single transit events.
The first is that the transit probability for long-period planets is very low;
scaling as $\propto P^{-5/3}$ for orbital periods longer than the
baseline of contiguous observations.
Therefore, even if long-period planets are intrinsically common, they will
still be underrepresented in a transiting sample.
The second challenge is that there are many signals in the observed
light curves caused by stochastic processes~--~both instrumental
(pointing jitter, temperature variations, \etc) and astrophysical (stellar
variability, \etc)~--~that can masquerade as transits.
Even using the most sophisticated methods for removing this variability, false
signal far outnumber the true single transits.

@@

To study the occurrence rate of long-period exoplanets, we have developed a
fully-automated method to search the archival \kepler\ light curves and
reliably discover long-period planets with only one or two transits.
Using this method, we search the 40,000 brightest Sun-like stars and detect
XXX long-period astrophysical transit candidates.
Of these, YYY are likely to be exoplanets with radii equal to or smaller than
Jupiter's radius.
Using transit injection and recovery, we measure the completeness of our
search procedure.
We estimate that the population level astrophysical false positive rate in
this candidate sample is about ZZZ\%.
By injecting synthetic transit signals into the \kepler\ light curves, we
measure the completeness of our search method.
Based on this catalog of exoplanet candidates and the search completeness, we
constrain the occurrence rate of long-period exoplanets.
For large planets, these results are consistent with previous measurements of
long-period, giant planet occurrence rates based on radial velocity surveys
and we demonstrate that small~--~and possibly rocky~--~planets are more common
than their larger counterparts even at long orbital periods.

@@

The model comparison step (step 2) is the key component of our method that
enables robust automation but it is also computationally expensive because we
must estimate the marginalized likelihoods for the light curve under several
different models describing a transit and many other processes that ``look''
like transits but are actually caused by noise.
This step is also conservative; unless a signal is a very convincing transit,
it won't pass the test.
In practice, this means that all but the highest signal-to-noise events will
be rejected at this step.
Therefore, in the inexpensive first step~--~the initial candidate search~--~we
can restrict the candidate list to high signal-to-noise events without a
substantial loss in detection efficiency.

\subsection{Initial candidate events}

It is not computationally feasible to run a full model comparison at every
time in the light curve so we must first find potentially interesting events.
For our purposes, ``interesting'' means high signal-to-noise and previously
unknown.

@@

As previously mentioned, the following steps will only detect high
signal-to-noise transits.
To avoid edge effects, we apodize this detection scalar near any large gaps in
the time series using a logistic function with width equal to one transit
duration.
Finally, we estimate the noise in the time series using a robust running
windowed variance estimate of the detection scalar and accepting peaks more
than 25-times this noise as candidates.

@@

In this step of the method, the goal is to discard any signals that are not
sufficiently ``transit-like'' in shape.
To quantify this decision, we perform a model comparison between a physical
transit model and a set of other parameterized models for systematics.
In order for a candidate to pass this vetting step, the transit model must be
``preferred'' to any other model as measured using the Bayesian Information
Criterion (BIC).
The BIC is not the optimal choice for this model comparison but it is
computationally intractable to compute thousands of precise marginalized
likelihoods for each model.
The BIC can be efficiently computed and it exhibits the desired
behavior~--~increasing with the likelihood but flexible models are
penalized~--~and we find that it performs sufficiently well in practice.

@@

For each model, we describe the data using a Gaussian Process (GP;
\dfmtodo{cite}) with a Mat\'ern-3/2 covariance and mean given by the chosen
model $m_k(t;\,\meanpars)$ parameterized by the parameter vector \meanpars.

@@

The goal of this project is to discover the transits of long-period planets
that have not yet been discovered.
Therefore, when studying the light curve of an eclipsing binary star or a star
with known transiting planet candidates~--~on shorter periods~--~we also remove
all the in-transit data for the candidate using the parameters provided by the
\project{NASA Exoplanet
Archive}\footnote{\url{http://exoplanetarchive.ipac.caltech.edu/}; We
downloaded the \texttt{cumulative} table of \kepler\ Objects of Interest on
2015-03-25.}.

@@

To measure the completeness of our long-period transit search method, we
exploit the fact that transit signals are sparse and rare.
Therefore, most light curves contain no transits and we can reliably measure
the recovery rate of our method on synthetic transit signals~--~with known
properties~--~injected into real light curves.
This procedure is standard practice in the transit literature and it has been
used to determine the completeness of the KOI catalog (\dfmtodo{cite}) and
other independent transit searches (\dfmtodo{cite}).

@@

The discovery and characterization of transiting planets based on a single
transit event is crucial for the future of transiting exoplanet surveys.
Many of the most dynamically influential planets~--~like Jupiter in our Solar
system~--~exhibit only a single transit in the full observational baseline of
the \kepler.
This will become even more of a problem as upcoming surveys move to shorter
contiguous observations.
For example, the \tess\ Mission is planned to get full-sky coverage at
half-hour cadence but most of the sky will only be targeted for a month.
This means that even habitable zone planets orbiting cool stars will transit
their host \emph{at most once in the entire lifetime of the Mission}!

@@

\section{Gaussian process regression}

Gaussian Processes (GPs) are a class of non-parametric, stochastic models that
have been demonstrated to be good effective models for the variability in
\kepler\ light curves.
A simple GP model can be used to capture residual non-transit variability in
light curves.
In this \paper, we use a GP model for two steps: light curve--level transit
shape vetting and parameter estimation.
A full discussion of GPs is beyond the scope of this \paper, so we will only
summarize the most relevant points here and direct an interested reader to
\dfmtodo{cite RW} for more details.

A GP model is specified by the following likelihood function
\begin{eqnarray}\eqlabel{gplike}
\mathcal{L} = \ln p(\bvec{y}\,|\,\meanpars,\,\kernpars) &=&
- \frac{1}{2}\,\bvec{r}(\meanpars)^\T\,K(\kernpars)^{-1}\,
    \bvec{r}(\meanpars)
- \frac{1}{2}\log\det K(\kernpars) - \frac{N}{2} \log{2\,\pi}
\end{eqnarray}
where \bvec{y} is a list of measurements in a scalar time series~--~in this
case, fluxes~--~measured at the times \bvec{t}, and
\begin{eqnarray}
\bvec{r}(\meanpars) &=& \bvec{y} - m(\bvec{t};\,\meanpars)
\end{eqnarray}
is the vector of residuals away from the mean model $m(\bvec{t};\,\meanpars)$.
For the purposes of this paper, we model the covariance matrix $K(\kernpars)$
using the Mat\'ern-3/2 kernel.
Under this model, the elements of $K(\kernpars)$ are given by
\begin{eqnarray}\eqlabel{matern}
\left[ K(\kernpars) \right]_{ij} &=& \sigma_i^2\,\delta_{ij}
    + \alpha^2 \left[ 1+\frac{|t_i - t_j|}{\sqrt{3}\,\tau} \right]
      \exp \left(-\frac{|t_i - t_j|}{\sqrt{3}\,\tau}\right)
\end{eqnarray}
where $\sigma_i$ is the reported uncertainty on the $i$-th measurement in the
time series and $\delta_{ij}$ is the Kronecker delta.

This covariance function (\eqalt{matern}) is specified by an amplitude
$\alpha$ and a time scale $\tau$ and we will simultaneously fit for these
hyperparameters $\kernpars=(\alpha,\,\tau)$ and the parameters of the mean
model \meanpars.
To efficiently find the parameter set that maximizes \eq{gplike} using a
non-linear optimization routine\footnote{We use the L-BFGS-B method as
implemented in SciPy \url{%
http://docs.scipy.org/doc/scipy/reference/generated/%
scipy.optimize.minimize.html}.},
it is useful to be able to compute the gradient of \eq{gplike} with respect to
the parameters \meanpars\ and \kernpars.
These gradients are given by
\begin{eqnarray}\eqlabel{gpmeangrad}
\frac{\dd\ln p(\bvec{y}\,|\,\meanpars,\,\kernpars)}{\dd \meanpars} &=&
\frac{\dd m(\bvec{t};\,\meanpars)}{\dd\meanpars}^\T \, K(\kernpars)^{-1} \,
    m(\bvec{t};\,\meanpars)
\end{eqnarray}
and
\begin{eqnarray}
\frac{\dd\ln p(\bvec{y}\,|\,\meanpars,\,\kernpars)}{\dd \kernpars} &=&
\frac{1}{2}\,\mathrm{Tr}\left(
    \left[ \bvec{\phi}\,\bvec{\phi}^\T - K(\kernpars)^{-1} \right]
    \,\frac{\dd K(\kernpars)}{\dd\kernpars}
\right)
\end{eqnarray}
where
\begin{eqnarray}
\bvec{\phi} &=& K(\kernpars)^{-1}\,\bvec{r}(\meanpars) \quad.
\end{eqnarray}

To evaluate \eq{gpmeangrad}, we must also evaluate the derivative of the mean
model with respect to its parameters.
For this \paper, the model $m(\bvec{t};\,\meanpars)$ will sometimes be the
limb-darkened, exposure time-integrated transit of a planet on a Keplerian
orbit and the parameters \meanpars\ are the physical and orbital parameters of
the planet and star.
Analytically computing the gradient of a simple transit model is possible
\citep{Pal:2008} but it becomes substantially more tedious as the model
becomes more realistic.
Therefore, we use a compile-time automatic differentiation library\footnote{%
More specifically, we use the \texttt{Jet} object from the BSD-licensed Ceres
Solver \url{http://ceres-solver.org}} \citep{Agarwal:2016} to compute first
derivatives of the full transit model to machine precision.
